Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-01 paper-amia-summit-2018/amia_summit_2018.tex
update data collection

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist} and a fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. \\

@@

As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm. Annotation column in Table~\ref{tab:anno_examp} shows the sequence of behavior codes from top to bottom, where counselor starts with an open-ended question and gets positive feedback at the end. It was noticed that same context represents different behavior codes which can be seen at row 7 and 9 in Table~\ref{tab:anno_examp}. Because human brain can remember the past behaviors, and classified the current context into different behavior codes after considering their past behaviors. In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, eickeler1998hidden, srivastava2007hmm, won2004training, chai2001folk}. However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variants of RNN models in our experiemnts: long short-term memory (LSTM) and gated recurrent unit (GRU).

\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property). Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUP-W, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.

@@

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation. Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews. \\

>>>>

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from the transcripts of 106 motivational interviews, which include a total of 41,764 segmented and annotated utterances. Each transcript consists of a counselor-adolescent and a counselor-caregiver session. Since the ultimate goal of a motivational interview is to increase the desire and ability of adolescents for a targeted behavior change, we considered only communication sequences from counselor-adolescent sessions and disregarded all communication sequences from counselor-caregiver sessions. The utterances were annotated based on MYSCOPE codebook \cite{carcone2013provider}, which are grouped into the adolescent and counselor code groups. Utterances were divided into successful and unsuccessful communication sequences. Successful communication sequences result in positive change talk and commitment language (a special class of change talk) statements by an adolescent, while unsuccessful sequences are the ones that result in negative change talk or commitment language and the sequences, in which no change talk or commitment language statements occur. Out of 4169 observed sequences, 3427 were positive and 742 were negative. For each of the probabilistic models (Markov chain and HMM), one model was trained using successful sequences and one model was trained using unsuccessful sequences. Statistics of the experimental dataset are presented in Table~\ref{tab:data_dist}. \\

@@

As can be seen from Table~\ref{tab:data_dist}, our data set is imbalanced and predominately composed of successful sequences with only a small percentage (17.80\%) of unsuccessful sequences. Usually, predictive accuracy misrepresents the performance of an employed algorithm for an imbalanced data. Because simply predict the majority class (successful) would provide a predictive accuracy 82.20\%. Therefore, it is important to handle imbalance data properly. Many solutions are proposed to deal with the imbalanced data sets, which can be divided into two categories: data and algorithmic levels. In this study, we utilized under and over sampling methods at data processing levels for balancing the adolescent obesity data set. Synthetic Minority Over-sampling (SMOTE) method is the most widely used method for over-sampling an imbalanced data set, in which new synthetic examples were generated from minority class \cite{chawla2002smote}. In this study, we generated synthetic examples along the borderline between minority examples and their selected nearest neighbors \cite{nguyen2011borderline}. On the other hand, under sampling method under samples the majority class by replacing a cluster of majority samples by the cluster centroid of a K-Means algorithm.  A fragment of an adolescent session transcript is presented in Table~\ref{tab:anno_examp}. Hence, annotation column shows the successful sequence of behavior codes from top to bottom, in particular, ($SS\rightarrow OQO\rightarrow HUPO\rightarrow OQTBN\rightarrow CHT+$) is a successful behavior sequence, where counselor starts with a structured session and gets positive feedback at the end. It was noticed that same context ``Yeah" and ``Yes'' represent different behavior codes CHT+ and HUPW, respectively. Because human brain can remember the past behaviors, and classify the current context into different behavior codes after considering their past behaviors. In this paper, we only consider those models that can capture the past events for the classification of behavior sequences.\\

@@

Our proposed baseline prediction method consists of two steps. In the first step, we model successful and unsuccessful patient-provider interactions using first- and second-order Markov Chain and Hidden Markov Model, which are popular probabilistic models for discrete observation sequences with finite vocabulary. In the second step, we classified each test sequence based on the maximum likelihood of generating that sequence from each model. Although HMM was originally developed for speech recognition \cite{rabiner1989tutorial}, it is one of the most widely used methods for sequence modeling \cite{mutsam2016maximum, won2004training, chai2001folk}. However, the latest advances in deep learning technologies show that deep learning, in particular, RNNs provide better results than conventional machine learning methods for the task of sequence classification. Therefore, we employed two special variants of RNN models in our experiemnts: long short-term memory (LSTM) and gated recurrent unit (GRU).

\textbf {Markov Chain (MC)} is a stochastic model for randomly changing systems, which assumes that the next state of a system only depends on its current state and not on its past states (Markov property). Generally, this assumption enables reasoning and computation with the models that would otherwise be intractable. For the sequential analysis, we built two Markov models $M$ and $\overline{M}$ summarizing provider strategies and patient responses in case of successful ($M$) and unsuccessful ($\overline{M}$) motivational interviews. A Markov model $M$ can be represented as a weighted directed graph $G = (V, E, p)$, in which:
\begin{itemize}
\item $V = \{CML+, CHT+, CHT-, AMB-, BLT, LUP+, LUP-, HUPW, ...\}$ is a set of vertices, consisting of adolescent and counselor MI behavior codes;
\item $E \subseteq V \times V$ is a set of edges corresponding to possible transitions from one MI behavior code to the other in a sequence;
\item $p_M:E\rightarrow[0...1]$ is a function that assigns probability $p(c_i|c_j)$ to an edge between the MI behavior codes $c_i$ and $c_j$ based on maximum likelihood estimator:
\begin{equation}
P_M(c_j|c_i) = \frac{n_{c_i,c_j}}{n_{c_i}}
\end{equation}
\end{itemize}
where $n_{c_i,c_j}$ and $n_{c_i}$ are the number of times a transition between the MI behavior codes $c_i$ and $c_j$ and the code $c_i$ has been observed in the training data. Given a Markov model $M$ (such that $S\subseteq V$), the probability that a sequence of MI behavior codes $S = \{C_1,...,C_N\}$ has been generated from a Markov model $M$ is:
\begin{equation}
P_M(S) = \prod_{i=2}^N p_M(c_i|c_1,\dots,c_{i-1})=\prod_{i=2}^N p_M(c_i|c_{i-1})
\end{equation}
In the second step, we quantify the likelihood of success of a given motivational interview at a certain time point given a sequence of MI behavior codes $S$ observed prior to that point using the following formula:
\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}
If $p(S\rightarrow CML+) > 0 $, the interview is predicted to result in positive change talk or commitment language.

@@

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, providing information can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when provided with general information that undermines their motivation. Analyzing such cases will allow the counselors to determine the general information that can be provided during the interviews. \\
